Credit Risk Distribution and Capital Savings

### 2020-03-16 Analysis/Draft/Draft.Rmd
Update paper draft

---
title: "Credit risk distribution and capital savings"
author:
- name: Lucas Avezum
  affiliation: CentER and EBC, Tilburg University; l.avezum@uvt.nl; +31134664531;  Warandelaan 2, 5037 AB Tilburg, NL, The Netherlands 
date: 
header-includes:
   - \usepackage{rotating, graphicx, dcolumn, longtable}
output:
  pdf_document:
    citation_package: natbib
    latex_engine: pdflatex
    template: tex-ms.tex
    fig_caption: yes
  html_document:
    df_print: paged
  word_document: default
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
anonymous: no
fontsize: 11pt
geometry: margin= 1in
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Regulatory arbitrage
JEL: G21, G28, G11
linkcolor: black
biblio-style: apsr
subtitle: ''
thanks: Very preliminary draft
abstract: "The Basel II accord introduced a concave mapping from probabilities of default (PD) to capital requiremets for banks adopting the internal ratings-based (IRB) approach. As a consequence, total capital requirement depends not only on the average PD but on the shape of the PD distribution. Using a novel hand-collected dataset, I propose a measure that disantangles the contribution of the PD distribution shape from the average PD effect to capital requirements. I document that the latter is an important factor of risk-weight variability across banks. I further investigate the prior ambiguous direction of its relationship with the business cycle and find it to be countercyclical. Altogether, my results show that issues related to RW manipulation and procyclicality of the IRB framework are not as severe as the literature suggests since the effect of the PD distribution shape was not taken into account."
toc: no
urlcolor: null
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
#library(relaimpo)            ## Calculate relative importance measures

@@

Bank capital adequacy requirements have evolved substancially since the first international standard, the 1988 Basel Accord. Arguably, at the center of these changes is the link of capital charges to asset risk [@basel2004international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced with the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models are the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit]. However, these same features may have adverse consequences. First, linking capital charges to asset risk exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. Second, the inherent flexibility of the internal ratings-based (IRB) approach may allow for differences in capital requirements that do not reflect portfolio risk [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits; @ferribank]. This paper contributes to the understanding of cyclicality and unintended variability of capital requirements by exploring the consequences of a feature of the IRB framework, which, to the best of my knowledge, I am the first to identify.

Under the IRB approach banks supply their own risk estimates to regulatory formulas which determine the amount of capital banks have to hold for each asset. Importantly, the mapping from probability of default (PD) and capital requirements is concave and capital requirements are determined on the asset level with total requirement being simply the sum of assets' individual requirement. This means, by Jensen’s inequality, that the sum of capital charges for two assets is lower than the charge for their average (in terms of PD). More generally, total capital requirement depends not only on banks' portfolio average PD but also on the shape of the PD distribution. For instance, given an average PD, more disperse portfolios yield lower capital charges. I call the difference in capital requirements stemming from differences in the shape of the portfolio PD distribution, while holding the location parameter constant, the PD distribution effect.

The first contribution of this paper is to introduce a dataset and method to quantify the PD distribution effect. The method is straightforward: compare the actual capital requirement, which depend on the PD distribution, to a counterfactual case where the requirement for the entire portfolio is calculated using the average PD. *Capital savings* is the difference between the two values and it measures the intensity of the PD distribution effect.

To calculate the *Capital savings* measure, one needs information on the distribution of banks' regulatory risk parameters. However, data at such granularity is not available in the common balance sheet datasets, and although very detailed, datasets from credit registers are restricted to one country, and may still lack the relevant information. Therefore, to overcome data limitations, I hand-collected information on the distribution of credit risk parameters from banks' Pillar-3 reports. The sample consists of 49 large banking groups that have adopted the IRB approach. The dataset covers 10 countries and goes from 2007 to 2018.

With this unique dataset, I start the analysis by investigating the contribution of the PD distribution effect to banks' overall risk-weight (RW). Several studies document significant RW variability across otherwise similar banks. Some take this evidence as indicative of either excessive subjectivity in the current capital requirement rules or risk measurement manipulation by banks to decrease capital charges [see, @mariathasan2014manipulation; @ferribank]. Others are more cautious and emphasize that part of RW variability is desirable since it reflects differences in risk among banks [@cannata2012inside; @BCBS2016regulatory; @EBA2019result]. I contribute to the literature by showing that the PD distribution effect is an important factor of the RW variation across banks. By extending the decomposition method proposed by @cannata2012inside I find that the PD distribution effect accounts for 28% of RW variation across banks. Moreover, the inclusion of the PD distribution effect in the decomposition reduces the variation explained by the component most susceptible of manipulation by 17.7 percentages points. This result implies that previous studies in the literature are potentially overestimating subjectivity and manipulation of RWs.

I then extend the analysis to investigate the relationship between *Capital savings* and the business cycle. A consequence of the shape of the regulatory formula is that it affect the cyclicality of capital requirements. @behn2016procyclical document that IRB portfolios are more procyclical than portfolios under the standardized approach. During economic downturns credit risk increases which raises capital requirements under the IRB approach. This characterizes the IRB approach as procyclical since banks have to cut down lending exactly when borrowers need it the most. We can interpret this evidence as the result of shifts to the location parameter of the PD distribution. However, shocks to credit risk can also affect the shape of the distribution. For instance, riskier borrowers prior to the shock can be more adversely affected compared to safer borrowers causing the PD distribution to become more fat-tailed. If that is the case, the amount of *Capital savings* will be affected.

I find the PD distribution effect to be countercyclical. I estimate that a 2.1% GDP growth rate (one standard deviation) decreases *Capital savings* by 5.9% on average. In other words, banks have to increase capital by 5.9% to keep their capital ratio constant. The effect is economically significant as it represents a quarter of the *Capital savings* standard deviation in the sample. I further analyse the relationship of the PD distribution moments with GDP and find that during economic expasions: average, variance and kurtosis decrease while the skewness increases. This evidence is in line with the idea that the upper-tail (riskier borrowers) of the PD distribution is disproportionately affected by the business cycle.

Several cross‐sectional tests further strengthen our results. Increases in capital requirements are more problematic for less well capitalized banks, which consequently adjust their IRB loans more in response to the credit risk shock. This suggests that our findings would be less pronounced if banks kept larger buffers over the regulatory minimum. Furthermore, the IRB effect is more pronounced for less profitable firms, which are likely to be more affected by the credit risk shock, and for firms to which IRB banks have a large exposure. As robustness, I calculate the gini coefficient for the PD distribution as an alternative measure of the PD distribution effect and I find similar results.

The remainder of this paper is organized as follows. Section 2 describes the capital requirement formula, the capital savings measure, and the hypothesis concerning this measure. Section 3 describes the data collected, and analyses the relative importance of capital savings in explaining differences in risk-weights across banks and time. Section 4 explains the empirical strategy. Section 5 presents evidence on the anticyclical feature of capital savings while Section 6 shows evidence of banks using this feature strategically to reduce capital charges. Finally, section 7 concludes.

@@

In the IRB approach, capital requirements are derived from risk-weight formulas that take as input banks' own estimates of credit risk drivers. The basic regulatory formula derives from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement.

Equation \ref{eq1} is the basic capital requirement formulas:

\begin{equation}
\label{eq1}
\text{K}= \sum_{i}EAD_iLGD_i N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(PD_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg) ,
\end{equation}

where the summation is over the assets and $N$ is the standard normal cumulative distribution. Basel accords segment credit portfolio into several categories, each having different additional features (maturity adjustments, risk-dependent correlation coefficient, among others) but without altering the basic structure shown in equation \ref{eq1}. The credit risk parameters estimated by banks, which are used as input on equation \ref{eq1}, are: Exposure at default (EAD), loss given default (LGD), and probability of default (PD).

@@

For the purpose of this paper, I focus on the mapping from PD to capital requirements. Figure \ref{fig:pdmap} shows that capital requirements is a concave function of PD. This means, by Jensen’s inequality, that the sum of capital charges for two assets is lower than the charge for their average (in terms of PD). More generally, the distribution of credit risk matters to calculate capital requirements with, for instance, disperse portfolios yielding relatively lower capital charges.

```{r parameters and data, echo=FALSE, include=FALSE}
set.seed(123)
LGD <- 1
M   <- 2.5
grid  <- seq(0.0001, 1, by=0.00001) 
data <- mapping_retail(grid,1,0.15,0.999)
```

@@

```{r, fig.height = 3, fig.width = 4,results="hide", fig.cap = pd_map_cap, fig.align = 'center', echo=FALSE, message=FALSE}
p <- ggplot(data = subset(data,subset = PD <0.2),
            mapping = aes(x = PD, y=K)) 
p + geom_line(alpha = 1) + 
  labs(x = "PD", 
         y = "Capital requirement per unit of exposure") +
  theme_bw()

@@

The first contribution of this paper is a measure of capital requirement reduction due to the concavity of the regulatory formula. The method behind the measure is simple: compare actual capital requirement to a counterfactual case where the requirement for the entire portfolio is calculated using the average PD. In practice, I calculate differences in risk-weighted assets (RWA) rather than capital requirements because it is the former that is used to calculate capital ratios. I define the counterfactual risk-weighted assets as:

\begin{equation}
\label{eq2}
\text{RWA}^c= 12.5\sum_{i}EAD_iLGD_iN\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\overline{PD})+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg),
\end{equation}

where the probability of default parameter for all assets is replaced by the portfolio weighted average:

\begin{equation*}
\overline{PD}=\frac{\sum_iEAD_iLGD_iPD_i}{\sum_iEAD_iLGD_i}.
\end{equation*}

The *Capital savings* measure is the difference between equation \ref{eq1} (multiplied by 12.5 to obtain RWA) and equation \ref{eq2}:

@@

*Capital savings* ($\text{RWA}^s$) is the amount that banks saves in capital in monetary units. A value of zero for this measures means that there is no capital savings, and implies a portfolio with a degenerate PD distribution. Any PD distribution with positive variance will yield a positive value of *Capital savings*. To avoid size effects and dealing with currency conversion I also consider a second measure, the *Capital savings ratio*, defined as follows:

\begin{equation}
\label{eq4}
\text{RWA}^r=\frac{\text{RWA}^c}{\text{RWA}}.
\end{equation}

*Capital savings ratio* ($\text{RWA}^r$) is the rate at which a bank saves capital. For instance, a value of 0.5 means this portfolio would require 50% more capital in the abscense of the PD distribution effect to keep the same capital ratio. As with *Capital savings*, a value of zero means no capital savings, and any PD distribution with positive variance will result in a positive value.

@@

A consequence of the shape of the regulatory formula is that it affect the cyclicality of capital requirements. @behn2016procyclical document that IRB portfolios are more procyclical than portfolios under the standardized approach. During economic downturns credit risk increases which raises capital requirements under the IRB approach. This characterizes the IRB approach as procyclical since banks have to cut down lending exactly when borrowers need it the most. We can interpret this mechanism as shifts to the location parameter of the PD distribution. However, shocks to credit risk can also affect the shape of the distribution. For instance, riskier borrowers prior to the shock can be more adversely affected compared to safer borrowers, causing the PD distribution to become more fat-tailed. As a result, the amount of capital savings will be affected.

Whether capital savings are pro- or anticyclical is an empirical question, as the direction of the cyclicality depends on how the economic cycle affects the credit risk distribution. I use a simple example to illustrate this ambiguity. Consider a bank with the asset side of its balance sheet composed of \$100 of a safer asset and \$100 of a riskier asset and three states of the economy. Without loss of generality, assume $LGD=1$ and $R=0.15$. In the initial state, safer and riskier assets have a 0.1% and 0.2% probability of default, respectively. According to equation \ref{eq1}, the actual RWA of this bank is \$63.9. The weighted average of this portfolio in terms of $PD$ is 0.15% which yields a counterfactual RWA of \$64.7 according to equation \ref{eq2}. Hence, in the initial state capital savings equals \$0.8.

Consider now a credit risk shock. In state A, both safer and riskier assets are affected, having their probabilities of default raised to 0.2% and 0.3%, respectively. In state B, only the riskier asset is affected, now with a probability of default equal to 0.4%. Note that the average $PD$ in both states are equal at 0.25%. However, relative to the initial state, the capital savings measure for state A decreases to \$0.45 while it increases to \$4.35 in state B. The mechanism of this example is simple: relative to the initial state, in state A only the location parameter of the PD distribution changes (standard deviation is equal to 0.05% in both states) while in state B also other moments change (standard deviation increases to 0.15%). Note that in both state A and B total capital requirements increase as average credit risk is higher relative to the initial state in both states. This is the result of IRB procyclicality shown by @behn2016procyclical. However, due to the increased portfolio dispersion, capital savings are anticyclical in state B. Conversely, while higher-order moments remained constant in state A, the PD distribution moved to a less concave part of the regulatory function consequently reducing capital savings. Therefore, capital savings are procyclical in state A. Table \ref{tab1} summarizes the example.

@@

\newpage

>>>>

---
title: "Credit Risk Distribution and Capital Savings"
author:
- name: Lucas Avezum
  affiliation: CentER and EBC, Tilburg University; l.avezum@uvt.nl; +31134664531;  Warandelaan 2, 5037 AB Tilburg, NL, The Netherlands 
date: 
header-includes:
   - \usepackage{rotating, graphicx, dcolumn, longtable}
output:
  pdf_document:
    citation_package: natbib
    latex_engine: pdflatex
    template: tex-ms.tex
    fig_caption: true
  html_document:
    df_print: paged
  word_document: default
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
anonymous: false
fontsize: 11pt
spacing: "double"
geometry: margin= 1in
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Procyclicality
JEL: G21, G28, G11
linkcolor: black
biblio-style: apsr
subtitle: ''
thanks: Preliminary draft.  
abstract: "The Basel II accord introduced a concave mapping from probabilities of default (PD) to capital requirements for banks adopting the internal ratings-based (IRB) approach. Therefore, total capital requirements depends not only on the portfolio's average PD but also on the shape of its PD distribution. Using a hand-collected dataset, I propose a method that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. While a larger average component increases capital charges, a larger shape component generates capital savings. I document that these capital savings are important sources of risk-weight variability across banks. I further investigate the relationship of capital savings with the business cycle and find it to be countercyclical. The results suggest that a recalibration of the current IRB framework can significantly reduce the procyclicality of capital requirements."
urlcolor: black
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
library(textablr)
library(sandwich)
library(lmtest)
#library(relaimpo)            ## Calculate relative importance measures

@@

The regulation on bank capital requirements have evolved substantially since the first international standard, the 1988 Basel Accord. At the heart of these changes is the link of capital charges to asset risk [@basel2004international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced in the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models, known as the internal ratings-based (IRB) approach. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit].

However, both risk sensitivity and the use of internal models may have adverse consequences. First, linking capital charges to asset risks exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. As the overall credit quality detiorates during economic downturns, capital requirements tighten, forcing banks to cut down lending. Second, the inherent flexibility of the IRB approach may allow for differences in capital requirements that do not reflect portfolio risk but rather modelling choices [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits;@berg2017analysis; @ferribank].

This paper contributes to understanding the cyclicality and variability of capital requirements by identifying the effects of an unexplored feature of the IRB framework on capital requirements. Under the IRB approach, banks supply their own risk estimates, such as probability of default (PD), to formulas set by the regulation which determine the amount of capital banks have to hold. Two characteristics of the framework are important for this analysis. First, the mapping from PD estimates to capital requirements is a concave function. Second capital requirements are calculated for each asset separately and then added up to reach total capital charges. Together, these characteristics imply that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in equal proportions. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD. Because of the concavity of the IRB formula, capital requirements for the second portfolio are greater than the first. More generally, total capital requirements depend not only on banks' portfolio average PD but also on the shape of the PD distribution.

I propose a measure that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. The method is straightforward: for each bank-year combination, compare actual capital requirements to the counterfactual case where requirements for the entire portfolio are calculated using the portfolio’s average PD. I label the shape component *Capital savings* because it reduces total capital requirements.

To calculate measures of *Capital savings*, information on the distribution of banks' regulatory risk parameters is needed. I hand-collected information on the distribution of credit risk parameters from banks' Pillar-3 reports. As is common in the literature, I calculate *Capital savings* in terms of risk-weighted assets (RWA), which are simply capital requirements times 12.5, or risk-weights (RW), which are RWA divided by assets. The sample consists of 52 large banking groups that have adopted the IRB approach. The dataset covers 13 countries and goes from 2007 to 2018. Data at such granularity is not available in the common balance sheet datasets, and although very detailed, datasets from credit registers are restricted to one country, and may still lack the relevant information.

With this novel dataset, I start by documenting the relative importance of *Capital savings* to banks' overall RW. By extending the decomposition method proposed by @cannata2012inside, I find that *Capital savings* accounts for 27% of the RW variation across banks. I also find that *Capital savings* measures are negatively correlated to other RW components. This correlation implies that banks with higher average portfolio riskness benefit the most from *Capital savings*. Lastly, although banks benefited from higher *Capital savings* as they moved their portfolio from the standardized to the IRB approach (known as the roll-out effect), the intensity of *Capital savings* has, on average, decreased in time. That is, for the same average PD, capital requirements have increased from 2008 to 2018.

I then extend the analysis to investigate the relationship between *Capital savings* and the business cycle. Prior evidence suggests that capital requirements are procyclical: during economic downturns banks tend to deleverage instead of raising capital. However, as credit conditions deteriorate (and banks react to these new credit conditions) the PD distribution and, consequently, the amount of *Capital savings*, also change. A priori, the direction of the business cycle effect on *Capital savings* is ambiguous. At one extreme, if the entire portfolio is equally affected, then there is only a shift of the location parameter of the PD distribution. Because the PD-RW formula elasticity is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are more sensitive to credit shocks, then capital savings are countercyclical. The ambiguity arises because *Capital savings* can only increase during a recession if the shape of the PD distribution is affected during the cycle.

I find *Capital savings* to be countercyclical. I estimate that a 1.7% GDP growth rate (one standard deviation) decreases *Capital savings* by 0.83 percentage points on average. In other words, banks have to increase capital by 0.83 percentage points to keep their capital ratio constant. The effect is economically significant as it represents a quarter of the capital ratio standard deviation in the sample. In a capital ratio decomposition exercise, I also find that the the cyclicality of *Capital savings* reduces the procyclicality of capital ratios by 11.6%. The results are robust to different measures of the shape component, such as, *Capital savings* in terms of ratios, *Capital savings* in monetary units, and using Gini coefficients.

I further analyse the relationship of the PD distribution moments with GDP growth and find that during economic expasions average PD and PD variance decrease while the PD kurtosis and PD skewness increase. This evidence implies that distributions become more symmetric and with less extreme values during recessions, suggesting a flight to safety behaviour by banks. As the recession reduces their existing portfolio credit quality (evidenced by the negative coefficient of average PD) banks respond by reducing risk-taking.

This papers contributes to two strands of the banking literature. First, several studies document significant RW variability across otherwise similar banks. Some take this evidence as indicative of either excessive subjectivity in the current capital requirement rules or risk measurement manipulation by banks to decrease capital charges [see, @mariathasan2014manipulation; @ferribank]. Others are more cautious and emphasize that part of RW variability is desirable since it reflects differences in risk among banks [@cannata2012inside; @BCBS2016regulatory;@berg2017analysis; @EBA2019result]. I contribute to the literature by showing that *Capital savings* are an important source of the RW variation across banks.

Second, while the literature shows that capital requirements, especially for portfolios under the IRB approach, are on average procyclical [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical], I identify a countercyclical component of the current IRB framework. A better understanding of this countercyclical component can help policy makers designing a less procyclical regulation in the future.

The remainder of this paper is organized as follows. Section 2 describes the capital requirement formula, the *Capital savings* measure, and the hypothesis concerning this measure. Section 3 describes the collected data and analyses the relative importance of *Capital savings* in explaining differences in risk-weights across banks and time. Section 4 presents evidence on the countercyclical feature of *Capital savings*. Finally, section 5 concludes.

@@

In the IRB approach, capital requirements are derived from risk-weight formulas that take as input banks' own estimates of credit risk parameters. The basic regulatory formula derives from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement.

Equation \ref{eq1} is the basic risk-weight formula:

\begin{equation}
\label{eq1}
\text{RW}= 12.5\frac{\sum_{i}\text{EAD}_i\text{LGD}_i N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\text{PD}_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)}{\sum_{i}\text{EAD}_i} ,
\end{equation}

where $i$ is an asset in the portfolio and $N$ is the standard normal cumulative distribution. Equation \ref{eq1} calculates the average RW for the porftolio. The total amount of RWA for this portfolio is $\text{RW}\times \sum_i\text{EAD}_i$ and total capital requirements $\text{RWA}\times 8\%$.

Basel accords segment credit portfolio into several categories, each having different additional features (maturity adjustments, risk-dependent correlation coefficient, among others) while keeping the basic structure of equation \ref{eq1}. The credit risk parameters estimated by banks, which are used as input on equation \ref{eq1}, are: Exposure at default (EAD), loss given default (LGD), and probability of default (PD).

@@

Figure \ref{fig:pdmap} plots the mapping from PD to RW according to equation \ref{eq1}. For my analysis, the crucial characteristic is that this mapping is concave. The concavity of the PD-RW mapping together with the bottom-up method to aggregate capital requirements imply that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in equal proportions. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD. Because of the concavity of the IRB formula, by Jensen’s inequality, the capital requirement for the second portfolio is greater than the first. More generally, total capital requirements depend not only on banks' portfolio average PD but also on the shape of the PD distribution. Figure \ref{fig:pdmap} illustrates the comparison. $\text{RW}^a$ and $\text{RW}^c$ are the average RW for the first portfolio and the second portfolio, respectively. The difference between  $\text{RW}^c$ and $\text{RW}^a$ is the monetary amount of *Capital savings* per unit of EAD due to the concavity of the IRB formula.

```{r parameters curve, echo=FALSE, include=FALSE}
set.seed(123)
LGD <- 1
M   <- 2.5
grid  <- seq(0.0001, 1, by=0.00001) 
data <- mapping_retail(grid,1,0.15,0.999)
PD.l <- 0.01
PD.h <- 0.15
PD.a <- (PD.l+PD.h)/2
PD.h2 <- 0.17
PD.h3 <- 0.16
PD.l3 <- 0.02
PD.a2 <- (PD.l+PD.h2)/2
PD.i  <- 0.19
K.l  <- as.numeric(mapping_retail(PD.l,1,0.15,0.999)["K"])
K.h  <- as.numeric(mapping_retail(PD.h,1,0.15,0.999)["K"])
K.a  <- as.numeric(mapping_retail(PD.a,1,0.15,0.999)["K"])
K.l3 <- as.numeric(mapping_retail(PD.l3,1,0.15,0.999)["K"])
K.h2 <- as.numeric(mapping_retail(PD.h2,1,0.15,0.999)["K"])
K.h3 <- as.numeric(mapping_retail(PD.h3,1,0.15,0.999)["K"])
K.a2  <- as.numeric(mapping_retail(PD.a2,1,0.15,0.999)["K"])
K.i  <- as.numeric(mapping_retail(PD.i,1,0.15,0.999)["K"])
K.s  <- ((K.h-K.l)/(PD.h-PD.l))*PD.a+K.l-((K.h-K.l)/(PD.h-PD.l))*PD.l
K.s2 <- ((K.h2-K.l)/(PD.h2-PD.l))*PD.a2+K.l-((K.h2-K.l)/(PD.h2-PD.l))*PD.l
K.s3 <- ((K.h3-K.l3)/(PD.h3-PD.l3))*PD.a2+K.l3-((K.h3-K.l3)/(PD.h3-PD.l3))*PD.l3
aux.df <- data.frame(PD.l = PD.l, PD.h = PD.h, PD.a = PD.a, PD.l3 = PD.l3, PD.a2,
                     K.l = K.l, K.h = K.h, K.a = K.a, K.s = K.s, K.l3 = K.l3, K.h2 = K.h2, 
                     K.h3 = K.h3, K.s3 = K.s3, K.s2 = K.s2)
```

@@

```{r, fig.height = 3, fig.width = 4,results="hide", fig.cap = pd_map_cap, fig.align = 'center', echo=FALSE, message=FALSE}
p <- ggplot(data = subset(data,subset = PD <0.2),
            mapping = aes(x = PD, y=K)) 
p + geom_line(alpha = 1) + 
  geom_segment(aes(x = PD.l, y = 0, xend = PD.l, yend = K.l), data = aux.df, linetype=3)+
  geom_segment(aes(x = PD.h, y = 0, xend = PD.h, yend = K.h), data = aux.df, linetype=3)+
  geom_segment(aes(x = PD.l, y = K.l, xend = PD.h, yend = K.h), data = aux.df, linetype=1)+
  geom_segment(aes(x = PD.a, y = K.a, xend = PD.a, yend = 0), data = aux.df, linetype=3)+
  geom_segment(aes(x = PD.a, y = K.s, xend = 0, yend = K.s), data = aux.df, linetype=5)+
  geom_segment(aes(x = PD.a, y = K.a, xend = 0, yend = K.a), data = aux.df, linetype=5)+
  geom_segment(aes(x=PD.l, xend=PD.l, y=K.a, yend=K.s),
               arrow = arrow(length = unit(0.15, "cm"),type = "closed"),linejoin='mitre',size = 0.5)+

  labs(x = "PD", 
         y = "Risk-weight") +
  theme_bw()+
  theme(axis.ticks = element_blank(),
        axis.line = element_line(),
        axis.title.x = element_blank(),
        panel.grid = element_blank())+
  expand_limits(x = 0, y = 0)+
  scale_x_continuous(breaks=c(PD.l,PD.a,PD.h,PD.i),labels=c("Low PD","Mean PD","High PD","PD"),expand = c(0, 0))+
  scale_y_continuous("",breaks=c(K.s,K.a,K.i),labels=c(expression(paste(RW^{a})),expression(paste(RW^{c})),"RW"),expand = c(0, 0)) +
  annotate("text", x = (PD.l+PD.a)/3 , y = (K.s+K.a)/2, label = "Capital \n Savings", size = 3)

@@

I propose a method that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. The method is a generalization of the example in Figure \ref{fig:pdmap}: for each bank-year combination, compare the actual capital requirement to the counterfactual case where the requirement for the entire portfolio is calculated using the portfolio’s average PD.

I define the counterfactual RW as:

\begin{equation}
\label{eq2}
\text{RW}^c= 12.5\frac{\sum_{i}\text{EAD}_i\text{LGD}_iN\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\overline{\text{PD}})+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)}{\sum_i \text{EAD}_i},
\end{equation}
in which the PD parameter for all assets is replaced by the portfolio weighted average:

\begin{equation*}
\overline{\text{PD}}=\frac{\sum_i\text{EAD}_i\text{LGD}_i\text{PD}_i}{\sum_i\text{EAD}_i\text{LGD}_i}.
\end{equation*}

The main measure of *Capital savings* used in this paper is the ratio of equation \ref{eq2} to equation \ref{eq1} as follows:

\begin{equation}
\label{eq4}
\text{RW}^r=\frac{\text{RW}^c}{\text{RW}}.
\end{equation}

$\text{RW}^r$ is the rate at which a bank saves capital. For instance, a value of 0.5 means this portfolio would require 50% more capital in the abscense of *Capital savings* to keep the same capital ratio. A value of zero means that there are no *Capital savings* and implies a portfolio with a degenerate PD distribution. Any PD distribution with positive variance results in positive *Capital savings*.

An alternative way to measure *Capital savings* is in monetary units as follows:

@@

@@

I then extend the analysis to investigate the relationship between *Capital savings* and the business cycle. Prior evidence suggests that capital requirements are procyclical: during economic downturns banks tend to deleverage instead of raising capital. However, as credit conditions deteriorates (and banks react to these new credit conditions) the PD distribution and, consequently, the amount of *Capital savings* also change. A priori the direction of the economic cycle effect on *Capital savings* is ambiguous. At one extreme, if the entire portfolio is equally affected, then there is only a shift of the location parameter of the PD distribution. Because the PD-RW formula elasticity is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are more sensitive to credit shocks, then capital savings are countercyclical. The ambiguity arises because *Capital savings* can only increase during a recession if the shape of the PD distribution is affect during the cycle.

@@

I use a simple example to illustrate this ambiguity. Consider a bank with the asset side of its balance sheet composed of \$100 of a safer asset and \$100 of a riskier asset and three states of the economy. Without loss of generality, assume $\text{LGD}=1$ and $\text{R}=0.15$. In the initial state, safer and riskier assets have a 0.1% and 0.2% probability of default, respectively. According to equation \ref{eq1}, the actual RWA of this bank is \$63.9. The weighted average of this portfolio in terms of $\text{PD}$ is 0.15% which yields a counterfactual RWA of \$64.7 according to equation \ref{eq2}. Hence, in the initial state *Capital savings* equal \$0.8.

```{r parameters curve 2, echo=FALSE, include=FALSE}
set.seed(123)
LGD <- 1
M   <- 2.5
grid  <- seq(0.0001, 1, by=0.00001) 
data <- mapping_retail(grid,1,0.15,0.999)
PD.l <- 0.01
PD.h <- 0.13
PD.a <- (PD.l+PD.h)/2
PD.h2 <- 0.19
PD.h3 <- 0.16
PD.l3 <- 0.04
PD.a2 <- (PD.l+PD.h2)/2
K.l  <- as.numeric(mapping_retail(PD.l,1,0.15,0.999)["K"])
K.h  <- as.numeric(mapping_retail(PD.h,1,0.15,0.999)["K"])
K.a  <- as.numeric(mapping_retail(PD.a,1,0.15,0.999)["K"])
K.l3 <- as.numeric(mapping_retail(PD.l3,1,0.15,0.999)["K"])
K.h2 <- as.numeric(mapping_retail(PD.h2,1,0.15,0.999)["K"])
K.h3 <- as.numeric(mapping_retail(PD.h3,1,0.15,0.999)["K"])
K.a2  <- as.numeric(mapping_retail(PD.a2,1,0.15,0.999)["K"])
K.s  <- ((K.h-K.l)/(PD.h-PD.l))*PD.a+K.l-((K.h-K.l)/(PD.h-PD.l))*PD.l
K.s2 <- ((K.h2-K.l)/(PD.h2-PD.l))*PD.a2+K.l-((K.h2-K.l)/(PD.h2-PD.l))*PD.l
K.s3 <- ((K.h3-K.l3)/(PD.h3-PD.l3))*PD.a2+K.l3-((K.h3-K.l3)/(PD.h3-PD.l3))*PD.l3
aux.df <- data.frame(PD.l = PD.l, PD.h = PD.h, PD.a = PD.a, PD.l3 = PD.l3, PD.a2,
                     K.l = K.l, K.h = K.h, K.a = K.a, K.s = K.s, K.l3 = K.l3, K.h2 = K.h2, 
                     K.h3 = K.h3, K.s3 = K.s3, K.s2 = K.s2)
```

```{r}
cyclicality_cap = "\\label{fig:pdmapcases}The figure shows the two possible outcomes of a negative credit risk shock on a portfolio. The left-hand side plot shows the case where capital savings are procyclical. The right-hand side plot shows the case where capital savings are countercyclical. "
```

```{r, cyclicality, fig.height = 3, fig.width = 6,results="hide",out.width = "90%", fig.align = 'center', echo=FALSE, message=FALSE, fig.cap= cyclicality_cap}
